---
title: "Choosing an independence rule"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing an independence rule}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(camtrapEvents)
```

## What the filter can and cannot infer

An independence filter assigns repeated photographs to analytical events. It
does not observe the true number of biological encounters. A time gap, a larger
visible group, a new age or sex class, and a changed identity label can all be
evidence for a new encounter, but each can also arise within one encounter.

The defensible workflow is therefore to state the rule, inspect sensitivity,
and validate a subset against evidence that was not itself used by the rule.
Unique markings, tags, synchronized video, or a blinded review of the original
images can provide such evidence. Reusing the same count and demographic fields
as both the rule input and its validation target is circular.

## Prepare one detection per row

Each input row must represent one unique photograph-species detection. Some
image-processing exports store male and female annotations for the same image on
separate rows. Combine those annotations before filtering. If a photograph ID
is available, ask the function to verify uniqueness:

```{r duplicate-example, error=TRUE}
d <- data.frame(
  station = "A",
  species = "Kobus kob",
  photo_id = c("IMG001", "IMG001"),
  datetime = as.POSIXct(c("2021-01-01 08:00:00", "2021-01-01 08:00:00"),
                        tz = "UTC"),
  females = c(2, 0),
  males = c(0, 1)
)

independent_events(
  d, "datetime", "station", "species",
  record_id = "photo_id", rule = "time_only"
)
```

The error is intentional. Whether split rows should be summed, maximized, or
otherwise reconciled depends on how the source software encoded annotations.
The package does not guess.

## Rule meanings

`time_only` starts a new event after a gap greater than `threshold`. Inside a
time burst, all records remain in one event. Use it when metadata are absent or
too uncertain to support classification.

`running_max` also starts an event when a numeric field exceeds its maximum in
the current burst. In a sequence of visible counts 5, 3, 5, the second and third
records do not open events. A later count of 6 can. The rule reduces repeated
splitting caused by oscillating counts, but improved visibility can still mimic
an arrival.

`any_change` starts an event whenever a selected field differs from the
preceding record. It supports categorical fields, including high-confidence
identity labels. With ordinary frame-level counts, however, it can repeatedly
split one moving group.

```{r rules}
recs <- data.frame(
  station = "A",
  species = "Kobus kob",
  datetime = as.POSIXct("2021-01-01 08:00:00", tz = "UTC") +
    60 * c(0, 5, 10, 15, 90),
  females = c(3, 5, 2, 3, 1),
  juveniles = c(0, 0, 1, 1, 0)
)

vapply(c("time_only", "running_max", "any_change"), function(r) {
  sum(independent_events(
    recs, "datetime", "station", "species",
    threshold = 30,
    rule = r,
    metadata = if (r == "time_only") NULL else c("females", "juveniles")
  )$independent)
}, numeric(1))
```

More retained events do not imply greater accuracy. The direction of error
depends on the balance between separate encounters that collide inside the
threshold and within-encounter metadata changes.

## Optional inner settling window

`metadata_refractory` supplies a second time scale. Records inside the short
window update the running maximum but cannot open a metadata-triggered event.
The outer threshold continues to define a new burst after inactivity.

```{r refractory}
rapid <- data.frame(
  station = "A",
  species = "Kobus kob",
  datetime = as.POSIXct("2021-01-01 08:00:00", tz = "UTC") +
    60 * c(0, 0.5, 1, 5),
  adults = c(3, 5, 2, 6)
)

independent_events(
  rapid, "datetime", "station", "species",
  threshold = 30,
  rule = "running_max",
  metadata = "adults",
  metadata_refractory = 2
)[, c("datetime", "adults", "independent", "event_id")]
```

The first three photographs remain in one event while establishing a maximum
of five. The later count of six occurs outside the two-minute settling window
and can open an event. This option trades sensitivity to very rapid true
arrivals for resistance to rapid visibility changes. Treat its value as a
parameter requiring ecological justification, not as a universal default.

## Count increments are not identities

`count_increment` divides the maximum observed count in a time burst across any
events created inside that burst. It avoids summing the full visible group size
again after every split. Its sum is a burst-wise observed maximum, not a count
of distinct animals. The legacy name `n_new` is retained only for compatibility.

## Sensitivity and reporting

```{r sensitivity}
data(waterhole)
s <- independence_sensitivity(
  waterhole, "datetime", "station", "species",
  thresholds = c(15, 30, 60, 120),
  metadata = c("males", "females", "juveniles"),
  count = "group_size"
)
head(s$overall)
```

At minimum, report:

1. the row unit and any consolidation of duplicate annotations;
2. the outer threshold and whether it is measured from the previous record or
   the previous retained event;
3. the metadata rule and fields used;
4. `min_increase` and `metadata_refractory` when non-default;
5. event totals under plausible alternative settings; and
6. how any manual validation was kept independent of the rule inputs.
